The Galactic Delay-Time Distribution of Type Ia Supernovae:\\ A Chemical Evolution Perspective

### 2023-06-14 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{aastex631}

@@

% Author list
\author[0000-0003-3781-0747]{Liam O. Dubay}
\author[0000-0001-7258-1834]{Jennifer A. Johnson}
\author[0000-0002-6534-8783]{James W. Johnson}
% \author[0000-0001-7775-7261]{David H. Weinberg}

@@

\begin{figure}
    \centering
    \includegraphics{figures/dtd_greggio05.pdf}
    \caption{Analytical DTDs from \citet[][solid curves]{Greggio2005-AnalyticalRates} and close approximation simple functions (dashed curves).}
    \label{fig:dtd-greggio05}
\end{figure}

@@

@@

\subsubsection{Accuracy of astroNN}
\label{sec:astronn-quality}

For $19784$ stars, or $\sim3.5\%$ of the sample, the astroNN [Fe/H] abundance estimate is $\gtrsim0.5$ dex lower than the value provided by ASPCAP. The vast majority of these stars are on the lower-main sequence according to \textit{Gaia} photometry, so we exclude stars with $\log(g) > 4$ in ASPCAP. In addition, 197 or $\sim3\%$ of stars with APOKASC-2 asteroseismic ages are reported to be $>5$ Gyr younger by astroNN. The APOKASC-2 age is unphysically high for many of these stars, often exceeding 15 Gyr. We identify prior mass loss as the cause of this discrepancy, so the astroNN ages are likely closer to the truth.

@@

@@

@@

@@

\begin{deluxetable*}{ll|cccc}
\label{tab:results}
\tablehead{
\colhead{DTD} & \colhead{SFH} & \colhead{MDF} & \colhead{ODF} & \colhead{[Fe/H]--[O/Fe]} & \colhead{Age--[O/Fe]}
}
\startdata
Power law               & Inside-out    & \meh  & \no   & \meh  & \no     \\
($\alpha=-1.1$)         & Late burst    & \meh  & \no   & \meh  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \meh    \\
                        & Two-infall    & \no   & \no   & \meh  & \no     \\
\hline
Power law               & Inside-out    & \meh  & \no   & \no   & \no     \\
($\alpha=-1.4$)         & Late burst    & \meh  & \no   & \no   & \no     \\
                        & Early burst   & \no   & \yes  & \no   & \meh    \\
                        & Two-infall    & \no   & \no   & \no   & \no     \\
\hline
Exponential             & Inside-out    & \yes  & \no   & \yes  & \no     \\
($\tau=1.5$ Gyr)        & Late burst    & \yes  & \no   & \yes  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \yes    \\
                        & Two-infall    & \yes  & \meh  & \meh  & \meh    \\
\hline
Exponential             & Inside-out    & \yes  & \meh  & \yes  & \meh    \\
($\tau=3$ Gyr)          & Late burst    & \yes  & \meh  & \yes  & \meh    \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\hline
Plateau                 & Inside-out    & \yes  & \meh  & \yes  & \no     \\
($W=0.3$ Gyr)           & Late burst    & \yes  & \no   & \yes  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \yes    \\
                        & Two-infall    & \yes  & \meh  & \yes  & \meh    \\
\hline
Plateau                 & Inside-out    & \meh  & \meh  & \yes  & \meh    \\
($W=1.0$ Gyr)           & Late burst    & \meh  & \meh  & \yes  & \meh    \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\hline
Prompt                  & Inside-out    & \meh  & \no   & \no   & \no     \\
($t_{50}=0.1$ Gyr)      & Late burst    & \meh  & \no   & \no   & \no     \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \yes  & \no   & \no   & \no     \\
\hline
Triple system           & Inside-out    & \yes  & \meh  & \meh  & \yes    \\
($t_{\rm{max}}=1$ Gyr)  & Late burst    & \yes  & \yes  & \meh  & \meh    \\
                        & Early burst   & \no   & \meh  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\enddata
\end{deluxetable*}

@@

>>>>

% Define document class
\documentclass[twocolumn,linenumbers,twocolappendix]{aastex631}

@@

% Author list
\author[0000-0003-3781-0747]{Liam O. Dubay}
\author[0000-0001-7258-1834]{Jennifer A. Johnson}
\author[0000-0002-6534-8783]{James W. Johnson}
\author[0000-0001-7775-7261]{David H. Weinberg}

@@

\begin{figure}
    \centering
    \includegraphics{figures/dtd_greggio05.pdf}
    \caption{Analytical DTDs from \citet[][solid curves]{Greggio2005-AnalyticalRates} and close approximation simple functions (dashed curves). Some functions are presented with a constant multiplicative factor for visual clarity.}
    \label{fig:dtd-greggio05}
    \script{dtd_greggio05.py}
\end{figure}

@@

\begin{enumerate}
    \item A simple power law with slope $\alpha$:
    \begin{equation}
        R(t) = A_\alpha (t/1\,\rm{Gyr})^\alpha
        \label{eq:powerlaw-dtd}
    \end{equation}
    where $A_\alpha = (\alpha+1)(t_{\rm max}^{\alpha+1} - t_{\rm min}^{\alpha+1})^{-1}$ is the normalization coefficient.
    We investigate the cases $\alpha=-1.1$ and $\alpha=-1.4$, following the field galaxy and galaxy cluster DTDs of \citet{Maoz2017-CosmicDTD}, respectively.

    \item A ``plateau'' model consisting of a flat slope of width $W$ followed by a power law decline with normalization $B_W$:
    \begin{equation}
        R(t) = 
        \begin{cases}
            B_W, & t < W \\
            B_W (t/W)^\alpha, & t \ge W
        \end{cases}
        \label{eq:plateau-dtd}
    \end{equation}
    where $B_W=[(W-t_{\rm min}) + \frac{t_{\rm max}^{\alpha+1} - W^{\alpha+1}}{(\alpha+1)W^\alpha}]^{-1}$ is the normalization coefficient.
    We investigate the cases $W=0.3$ Gyr and $W=1$ Gyr, taking $\alpha=-1.1$ for all plateau models. Figure \ref{fig:dtd-greggio05} illustrates the similarity between these models and two different treatments of the analytical double-degenerate DTD from \citet{Greggio2005-AnalyticalRates}.

    \item An exponentially declining DTD with timescale $\tau$ and normalization $C_\tau$:
    \begin{equation}
        R(t) = 1/\tau \exp(-t/\tau)
        \label{eq:exponential-dtd}
    \end{equation}
    We investigate the case where $\tau=1.5$ Gyr, which has been used in previous studies \citep[e.g.,][]{Schonrich2009-RadialMixing,Weinberg2017-ChemicalEquilibrium} and is similar to the analytical single-degenerate DTD from \citet{Greggio2005-AnalyticalRates}. We also investigate a longer timescale of $\tau=3$ Gyr.

    \item A DTD in which 50\% of SNe Ia belong to a ``prompt'' Gaussian component at small $t$ and the other 50\% form an exponential tail at large $t$, normalized by $D_{t_p,\sigma,\tau}$:
    \begin{equation}
        R(t) = D_{t_p,\sigma,\tau} \Big[\exp\Big(-\frac{(t-t_p)^2}{2\sigma^2}\Big) + \exp(-t/\tau)\Big]
        \label{eq:prompt-dtd}
    \end{equation}
    Following the two-population model from \citet{Mannucci2006-TwoPopulations}, we take $t_p=50$ Myr, $\sigma=15$ Myr, and $\tau=3$ Gyr, which results in roughly 50\% of SNe Ia exploding within $t<100$ Myr.

    \item A DTD based on simulations of triple-system evolution by \citet{Rajamuthukumar2022-TripleEvolution}. We approximate their DTD as a special case of the plateau model (Equation \ref{eq:plateau-dtd}) where the initial rate is quite low until an instantaneous rise to the plateau value at time $t_{\rm rise}$:
    \begin{equation}
        R(t) = 
        \begin{cases}
            \epsilon B_W, & t < t_{\rm rise} \\
            B_W, & t_{\rm rise} \leq t < W \\
            B_W (t / W) ^ \alpha, & t \geq W
        \end{cases}
        \label{eq:triple-dtd}
    \end{equation}
    We take $t_{\rm rise}=0.5$ Gyr, $W=0.5$ Gyr, $\alpha=-1.1$, and $\epsilon=0.05$ so the initial rate is 5\% of the peak rate.
\end{enumerate}

@@

We use APOGEE DR17 \citep{Abdurro'uf2022-SDSSIV-DR17}. Our selection criteria are presented in Table \ref{tab:sample-selection}, producing a final sample of 192,926 stars.

\begin{deluxetable*}{lll}
    \tablecaption{Sample Selection\label{tab:sample-selection}}
    \tablehead{
        \colhead{Parameter} & \colhead{Range} & \colhead{Notes}
    }
    \startdata
        $\log g$            & $1.0 < \log g < 3.8$          & Select giants only \\
        $T_{\rm eff}$       & $3500 < T_{\rm eff} < 5500$ K & Reliable temperature range \\
        $S/N$               & $S/N > 80$                    & Required for accurate stellar parameters \\
        ASPCAPFLAG Bits     & $\notin$ 23                   & Remove stars flagged as bad \\
        EXTRATARG Bits      & $\notin$ 0, 1, 2, 3, or 4     & Select main red star sample only \\
    \enddata
\end{deluxetable*}

\subsubsection{Stellar Ages}
\label{sec:stellar-ages}

191,179 stars in our sample have astroNN ages and 71,684 have ages from \citet{Leung2023-Ages}.

% For $19784$ stars, or $\sim3.5\%$ of the sample, the astroNN [Fe/H] abundance estimate is $\gtrsim0.5$ dex lower than the value provided by ASPCAP. The vast majority of these stars are on the lower-main sequence according to \textit{Gaia} photometry, so we exclude stars with $\log(g) > 4$ in ASPCAP. In addition, 197 or $\sim3\%$ of stars with APOKASC-2 asteroseismic ages are reported to be $>5$ Gyr younger by astroNN. The APOKASC-2 age is unphysically high for many of these stars, often exceeding 15 Gyr. We identify prior mass loss as the cause of this discrepancy, so the astroNN ages are likely closer to the truth.

@@

\subsubsection{Stellar Migration}
\label{sec:migration}

\citet{Johnson2021-Migration}

We fit a Gaussian to the distribution of $\Delta R = R_{\rm final} - R_{\rm initial}$ from the \texttt{h277} output in bins of $R_{\rm initial}$ and age. Each Gaussian is centered at 0 and we find that the scale $\sigma_{\Delta R}$ is best described by the function

\begin{equation}
    \sigma_{\Delta R} = 1.35\,{\rm kpc}(R_{\rm form}/8\,{\rm kpc})^{0.61} (\tau/1\,{\rm Gyr})^{0.33}
    \label{eq:radial-migration}
\end{equation}

\noindent where $\tau$ is the age of the star particle.

We fit a sech$^2$ function \citep{Spitzer1942} to the distribution of midplane distances $z_{\rm final}$. Vertical migration away from the midplane does not affect the chemical evolution simulation, but we do use $z_{\rm final}$ in our analysis. The probability density function (PDF) of $z_{\rm final}$ given some scale height $h_z$ is

\begin{equation}
    {\rm PDF}(z_{\rm final}) = \frac{1}{4 h_z} {\rm sech}^2\Big(\frac{z_{\rm final}}{2 h_z}\Big)
    \label{eq:sech-pdf}
\end{equation}

\noindent and the corresponding cumulative distribution function (CDF) is

\begin{equation}
    {\rm CDF}(z_{\rm final}) = \frac{1}{1 + \exp(-z_{\rm final} / h_z)}.
    \label{eq:sech-cdf}
\end{equation}

\noindent We fit the above function to the distributions of $z_{\rm final}$ in h277 in varying bins of $\tau$ and $R_{\rm final}$ and found that $h_z$ is best described by the function

\begin{equation}
    h_z = (0.25\,{\rm kpc}) 
    \exp\Big(\frac{\tau-5\,{\rm Gyr}}{7.0\,{\rm Gyr}}\Big)
    \exp\Big(\frac{R_{\rm final}-8\,{\rm kpc}}{6.0\,{\rm kpc}}\Big)
    \label{eq:scale-height}
\end{equation}

@@

A 150 Myr minimum delay time is incompatible with the prompt DTD model, which has $\sim 50$ of SNe Ia explode in the first 100 Myr.

@@

\begin{deluxetable}{lll}
    \tablecaption{Median Parameter Uncertainties\label{tab:uncertainties}}
    \tablehead{
        \colhead{Parameter} & \colhead{Median Uncertainty} & \colhead{Source}
    }
    \startdata
        [Fe/H] & $9.2\times10^{-3}$ & APOGEE \\
        $\rm [O/Fe]$ & $1.8\times10^{-2}$ & APOGEE \\
        log(Age/Gyr) & $0.12$ & \citet{Leung2023-Ages}
        % Age/Gyr & 29\% & astroNN
    \enddata
\end{deluxetable}

\subsection{Comparison to APOGEE MDFs}

To quantify the similarity between the MDFs generated by VICE and the observed APOGEE MDF, we compute the Kullback-Leibler (KL) divergence \citep{KullbackLeibler1951}, defined as

\begin{equation}
\label{eq:kl-divergence}
D_{\rm{KL}}(P \parallel Q) = \int_{-\infty}^{\infty} p(x) \log\Big(\frac{p(x)}{q(x)}\Big) dx
\end{equation}

\noindent for distributions $P$ and $Q$ with probability density functions $p(x)$ and $q(x)$. In this case, $P$ is the APOGEE MDF, $Q$ is the model MDF, and $x$ is [Fe/H]. For each model SFH and DTD, we compute $D_{\rm{KL}}$ in 18 different galactocentric zones defined by radial bins $3-5$ kpc, $5-7$ kpc, $7-9$ kpc, $9-11$ kpc, $11-13$ kpc, and $13-15$ kpc and absolute $z$-height bins $0-0.5$ kpc, $0.5-1$ kpc, and $1-2$ kpc. The KL divergence for the entire model is taken to be the average of $D_{\rm{KL}}$ for each zone, weighted by the number of APOGEE stars in each zone.

\subsection{Comparison to APOGEE ODFs}

@@

\begin{deluxetable*}{ll|cccc}
\label{tab:results}
\tablehead{
\colhead{DTD} & \colhead{SFH} & \colhead{MDF} & \colhead{ODF} & \colhead{[Fe/H]--[O/Fe]} & \colhead{Age--[O/Fe]}
}
\startdata
Power law               & Inside-out    & \meh  & \no   & \meh  & \no     \\
($\alpha=-1.1$)         & Late burst    & \meh  & \no   & \meh  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \meh    \\
                        & Two-infall    & \no   & \no   & \meh  & \no     \\
\hline
Power law               & Inside-out    & \meh  & \no   & \no   & \no     \\
($\alpha=-1.4$)         & Late burst    & \meh  & \no   & \no   & \no     \\
                        & Early burst   & \no   & \yes  & \no   & \meh    \\
                        & Two-infall    & \no   & \no   & \no   & \no     \\
\hline
Exponential             & Inside-out    & \yes  & \no   & \yes  & \no     \\
($\tau=1.5$ Gyr)        & Late burst    & \yes  & \no   & \yes  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \yes    \\
                        & Two-infall    & \yes  & \meh  & \meh  & \meh    \\
\hline
Exponential             & Inside-out    & \yes  & \meh  & \yes  & \meh    \\
($\tau=3$ Gyr)          & Late burst    & \yes  & \meh  & \yes  & \meh    \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\hline
Plateau                 & Inside-out    & \yes  & \meh  & \yes  & \no     \\
($W=0.3$ Gyr)           & Late burst    & \yes  & \no   & \yes  & \no     \\
                        & Early burst   & \no   & \yes  & \meh  & \yes    \\
                        & Two-infall    & \yes  & \meh  & \yes  & \meh    \\
\hline
Plateau                 & Inside-out    & \meh  & \meh  & \yes  & \meh    \\
($W=1.0$ Gyr)           & Late burst    & \meh  & \meh  & \yes  & \meh    \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\hline
Prompt                  & Inside-out    & \meh  & \no   & \no   & \no     \\
($t_{\rm max}=0.05$ Gyr)& Late burst    & \meh  & \no   & \no   & \no     \\
                        & Early burst   & \no   & \yes  & \no   & \yes    \\
                        & Two-infall    & \yes  & \no   & \no   & \no     \\
\hline
Triple system           & Inside-out    & \yes  & \meh  & \meh  & \yes    \\
($t_{\rm{max}}=1$ Gyr)  & Late burst    & \yes  & \yes  & \meh  & \meh    \\
                        & Early burst   & \no   & \meh  & \no   & \yes    \\
                        & Two-infall    & \meh  & \meh  & \meh  & \meh    \\
\enddata
\end{deluxetable*}

@@

Blah

\begin{acknowledgments}
Funding for the Sloan Digital Sky 
Survey IV has been provided by the 
Alfred P. Sloan Foundation, the U.S. 
Department of Energy Office of 
Science, and the Participating 
Institutions.

SDSS-IV acknowledges support and 
resources from the Center for High 
Performance Computing  at the 
University of Utah. The SDSS 
website is www.sdss4.org.

SDSS-IV is managed by the 
Astrophysical Research Consortium 
for the Participating Institutions 
of the SDSS Collaboration including 
the Brazilian Participation Group, 
the Carnegie Institution for Science, 
Carnegie Mellon University, Center for 
Astrophysics | Harvard \& 
Smithsonian, the Chilean Participation 
Group, the French Participation Group, 
Instituto de Astrof\'isica de 
Canarias, The Johns Hopkins 
University, Kavli Institute for the 
Physics and Mathematics of the 
Universe (IPMU) / University of 
Tokyo, the Korean Participation Group, 
Lawrence Berkeley National Laboratory, 
Leibniz Institut f\"ur Astrophysik 
Potsdam (AIP),  Max-Planck-Institut 
f\"ur Astronomie (MPIA Heidelberg), 
Max-Planck-Institut f\"ur 
Astrophysik (MPA Garching), 
Max-Planck-Institut f\"ur 
Extraterrestrische Physik (MPE), 
National Astronomical Observatories of 
China, New Mexico State University, 
New York University, University of 
Notre Dame, Observat\'ario 
Nacional / MCTI, The Ohio State 
University, Pennsylvania State 
University, Shanghai 
Astronomical Observatory, United 
Kingdom Participation Group, 
Universidad Nacional Aut\'onoma 
de M\'exico, University of Arizona, 
University of Colorado Boulder, 
University of Oxford, University of 
Portsmouth, University of Utah, 
University of Virginia, University 
of Washington, University of 
Wisconsin, Vanderbilt University, 
and Yale University.
\end{acknowledgments}
